---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on COVID Impact on Work
#### Bhargav Yellepeddi, Neel Rajan, Amaya Bayoumi, Ananya Ramji

## Introduction

In this analysis, we explore a dataset detailing information about how covid had an impact on people's work lives.

The goal of this project is to explore how Pandemics like COVID can impact various aspects of our lives. When the COVID-19 Pandemic occurred, society as a whole was not at all prepared on how to react, which caused lots of famine in many counties and sent the world into chaos. By analyzing this dataset, people can better understand what areas of life will be impacted the most, and thus learn how to better prepare for future events such as COVID. Ultimately, our goal is to determine the best methods that can minimize the impacts of Pandemic-like events.

To achieve this objective, we will explore the following questions:

1. What cleaning methods were applied, and why?

2. What is the distribution of the Productivity_Change Variable?

3. What is the impact of working from home on productivity change?
-   Stress Level & Working from Home
-   Sector
-   Sector & Working from Home
-   Childcare Responsibilities & Working from Home
-   Health Issues & Working from Home

4. Are there significant correlations between productivity_change and other numeric variables?

5. Do employees who experienced salary cuts show different productivity trends compared to those with stable salaries?

6. What was the impact of Childcare Responsibilities, Commuting Changes, and Health Issues on Technology Adaptation Levels

7. What are the key takeaways or recommendations based on our analysis?

8. What can we improve if this analysis were conducted again?

These are the primary questions we seek to answer in this final project. From the information drawn from these questions, we will be able to derive meaningful insights on the best practices to use during Pandmeics and better understand how to prepare for them.

## Data
### Structure
    The link to the data set is the following: https://www.kaggle.com/datasets/willianoliveiragibin/covid-19-on-working-professionals/data. The Kaggle website constains a CSV file containing all the data collected to detail the impact COVID had on people's work lives. There are 15 columns for 15 variables, where each variable describes a different aspect of work and whether or not it has had any change or effect from COVID. All of the data-points are neatly organized into one CSV file, where there are exactly 10,000 rows.

    One of the most important things to note about this dataset is that fact that most of the variables report their data in a binary format. In other words, most of the data is listed in 0's and 1's, where 0 conveys that the given data-point has not been effected or changed for its given variable/column, while 1 conveys that it has. Although this may seem subtle, this has lead to a variety of issues and mis-interpretations throughout this project, which have been since fixed of course. 

    Where normally graphs depict the amount of change a variable has had, binary-based variables will depict whether or not there was any change and how much of the data has changed. When it comes to this project specifically, this is actually a valid approach as all we need to see is what variables changed in response to COVID-19. The variables that had less change indicate that they are more resilient and are of less worry, whereas the variables that have a higher proportion of change are the ones that indicate the need for more caution and preparation.

    With that said, we now move on to viewing the dataset and gaining an overall synopsis of what to expect:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
library(ggplot2)
library(moments)
library(dplyr)
library(tidyr)

# Read the CSV file
data <- read_csv("covid_impact_on_work new.csv")
```

```{r}
# View the data 
head(data)
```

### Cleaning
#### Overview the Data
First, we will take an overview of the Data to see what aspects needed cleaning.

```{r}
colnames(data) # View the column names to ensure proper references
```

As shown, there are 15 variables/columns.

```{r}
str(data)      # Structure of the data
```

This code snippet reveals the structure of the Dataset. Some key things to note are that the binary variables are all listed as numeric data types while the rest are character data types.

```{r}
summary(data)  # Summary statistics
```

The summary statistics tells us quite a few things. One of the key pieces of information to take away is the proportion of data-points that were effected by their respective variables. This mainly applies to the binary/numeric variable types as the character types simply list that they are characters. The binary types, on the other hand, show their mean values, which represents the proportion of data that was effected in their column. For example, the "Job_Security" column lists a 0.4049 value for the mean, which means that around 40% of the people in the data set experinced and impact or change to their job securities during the impact. This indicates that job securtity was more stable than not for the average person during the pandemic. In contrast, "Team_Collaboration_Challenges" lists a mean value of 0.7006, meaning that 70% of people in the dataset experinced team collaboration challenges, indicating that this aspect is vulnerable to Pandmeic-like events. Overall, this information alone provides valuable insights and can allow society to better prepare for future events.

```{r}
colSums(is.na(data)) # Check missing values
```

As shown here, all the columns have a 0 for null values, which means that there are no empty values in any column. This means that we will not have to worry about accounting for null values during the data cleaning process.

### Variables

- Stress_Level: Indicates the employee's stress level, categorized into levels such as low, moderate, or high.

- Sector: Specifies the industry or sector (e.g., Retail, IT) in which the employee works.

- Increased_Work_Hours: A binary variable (1/0) indicating whether the employee experienced an increase in work hours.

- Work_From_Home: A binary variable (1/0) showing whether the employee is working from home.

- Hours_Worked_Per_Day: Captures the number of hours the employee works in a day, though the values in the sample data need cleaning.

- Meetings_Per_Day: Represents the number of meetings the employee attends per day, though concatenated values in the sample require cleaning.

- Productivity_Change: A binary variable (1/0) indicating whether the employee experienced a change in productivity.

- Health_Issue: A binary variable (1/0) denoting whether the employee reported any health issues.

- Job_Security: A binary variable (1/0) indicating whether the employee feels secure in their job.

- Childcare_Responsibilities: A binary variable (1/0) indicating if the employee has childcare responsibilities.

- Commuting_Changes: A binary variable (1/0) showing whether the employee's commute has changed.

- Technology_Adaptation: A binary variable (1/0) indicating if the employee had to adapt to new technology for work.

- Salary_Changes: A binary variable (1/0) representing whether the employee's salary changed.

- Team_Collaboration_Challenges: A binary variable (1/0) indicating if the employee faced challenges collaborating with their team.

- Affected_by_Covid: A binary variable (1/0) specifying whether the employee's work or life has been impacted by COVID-19.

Although we use most of these variables througout this analysis, our variables of focus are going to be the "Productivity_Change" and "Stress_Level" since these variables will tell us how society is doing as a whole as they reflect the state of the people.

## Results




## Conclusion

In conclusion, our analysis on the impact that COVID had on people's work lives has revealed how well prepared and un-prepared society is for pandemic-like events. We have also confirmed what works best and what doesn't when attempting to adapt to such events.