---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on COVID Impact on Work
#### Bhargav Yellepeddi, Neel Rajan, Amaya Bayoumi, Ananya Ramji

## Introduction

In this analysis, we explore a dataset detailing information about how covid had an impact on people's work lives.

The goal of this project is to explore how Pandemics like COVID can impact various aspects of our lives. When the COVID-19 Pandemic occurred, society as a whole was not at all prepared on how to react, which caused lots of famine in many counties and sent the world into chaos. By analyzing this dataset, people can better understand what areas of life will be impacted the most, and thus learn how to better prepare for future events such as COVID. Ultimately, our goal is to determine the best methods that can minimize the impacts of Pandemic-like events.

To achieve this objective, we will explore the following questions:

1. What cleaning methods were applied, and why?

2. What is the distribution of the Productivity_Change Variable?

3. What is the impact of working from home on productivity change?
-   Stress Level & Working from Home
-   Sector
-   Sector & Working from Home
-   Childcare Responsibilities & Working from Home
-   Health Issues & Working from Home

4. Are there significant correlations between productivity_change and other numeric variables?

5. Do employees who experienced salary cuts show different productivity trends compared to those with stable salaries?

6. What was the impact of Childcare Responsibilities, Commuting Changes, and Health Issues on Technology Adaptation Levels

7. What are the key takeaways or recommendations based on our analysis?

8. What can we improve if this analysis were conducted again?

These are the primary questuons we seek to answer in this final project. From the information drawn from these questions, we will be able to derive meaningful insights on the best practices to use during Pandmeics and better understand how to prepare for them.

## Data
### Structure
    The link to the data set is the following: . The Kaggle website constains a CSV file containing all the data collected to detail the impact COVID had on people's work lives. There are 15 columns for 15 variables, where each variable describes a different aspect of work and whether or not it has had any change or effect from COVID. All of the data-points are neatly organized into one CSV file, where there are exactly 10,000 rows.

    One of the most important things to note about this dataset is that fact that most of the variables report their data in a binary format. In other words, most of the data is listed in 0's and 1's, where 0 conveys that the given data-point has not been effected or changed for its given variable/column while 1 conveys that it has. Although this may seem subtle, this has lead to a variety of issues and mis-interpretations throughout this project, which have been since fixed of course. 

    Where normally graphs depict the amount of change a variable has had, binary-based variables will depict whether or not there was any change and how much of the data has changed. When it comes to this project specifically, this is actually a valid approach as all we need to see is what variables changed in response to COVID-19. The variables that had less change indicate that they are more resilient and are of less worry, whereas the variables that have a higher proportion of change are the ones that indicate the need for more caution and preparation.

    With that said, we now move on to viewing the Dataset and its structure:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
library(ggplot2)
library(moments)
library(dplyr)
library(tidyr)

# Read the CSV file
data <- read_csv("covid_impact_on_work new.csv")
```

```{r}
# View the data 
head(data)
```

### Cleaning
#### Overview the Data
First, we will take an overview of the Data to see what aspects needed cleaning.
```{r}
# Inspect the dataset
colnames(data) # View the column names to ensure proper references
str(data)      # Structure of the data
summary(data)  # Summary statistics
colSums(is.na(data)) # Check missing values
```


### Variables



## Results




## Conclusion